Dynamics of thin liquid film flows studied with the lattice Boltzmann method

### 2023-08-10 chapters/Theory.tex
All references polished, conclusion missing

\section{The Navier-Stokes equation}
\label{sec:navier_stokes_sec}
Mathematically the Navier-Stokes equation can be derived from one definition and two constraints.
The two constraints are mass and momentum conservation.
Mass conservation simply states that mass cannot be lost or created.
The same argument holds for the momentum.
By Noethers theorem every conserved quantity yields a symmetry and vice versa.
In this case conservation of momentum allows us to use the constraints on the solutions that translational symmetry requires.
Put simply a sliding drop experiment has to yield the same result in any laboratory on earths surface, independent of the exact location of the laboratory.
The last ingredient for the derivation is the definition of a fluid.
A fluid is ``\textit{a substance that flows and is not solid}`` according to the Cambridge Dictionary.
This definition can be refined, because a fluid is either a gas or a liquid that continuously deforms under shear stress, at least in the case of a Newtonian fluid.
And as such there is a linear relation between shear rate and shear velocity for a Newtonian fluid.
Clearly there are other liquids (which encloses granular systems to some extent) that exhibit different behaviour, for example toothpaste.
As long as there is no force applied to the tube the toothpaste does not flow.
Even the acceleration due gravity when pointing the tube towards the ground does not exert enough force to generate a flow.
In contrast to Newtonian liquids those liquids have a different response function to shear, they can be addressed as \textit{power law} fluids.
While the phenomenology of these fluids, e.g. shear thinning or thickening, is very interesting and important in many industrial applications it is beyond the scope of this work. 
That said, in Chapter~\ref{chapter:second_paper} an extension towards more complex flows is presented. 
Instead of however changing the shear response we model thermal fluctuations to approximate the stochastic thin film equation (STF).
In fact, the step to derive an extension to our approach are nicely summarized in Chapter~\ref{chapter:second_paper}. 
It should serve as a guideline for further features, such as viscoelasticity for non-Newtonian liquids~\cite{bouchutNewModelShallow2013}.

@@

For the momentum equation the same line of arguments is applied, however momentum can change for different reasons.
Momentum can for example be advected with the flow from or into the volume $V_0$.
Pressure on the other hand can be a source of momentum change and of course forces that are applied to the fluid are as well sources~\cite{krugerLatticeBoltzmannMethod2017}.
Collecting these terms and writing down an equation for the momentum we have,
\begin{equation}\label{eq:navier_stokes_1}
    \partial_t \int_{V_0}\rho\vec{u}\diff V = -\oint_{\partial V_0}(\rho\vec{u}\vec{u})\cdot\diff\vec{A} - \oint_{\partial V_0} p\diff\vec{A} + \int_{V_0} \vec{F}\diff V, 
\end{equation}
where $\vec{u}\vec{u}$ is used to denote the outer product
\begin{equation}
    \mathbf{u}\mathbf{u} = \begin{bmatrix}
    u_x u_x & u_x u_y & u_x u_z \\
    u_y u_x & u_y u_y & u_y u_z \\
    u_z u_x & u_z u_y & u_z u_z
    \end{bmatrix} .
\end{equation}
Using the divergence theorem again to reformulate everything in terms of volume integrals and dropping the integration the above equation becomes
\begin{equation}\label{eq:navier_stokes_2}
    \partial_t(\rho\vec{u}) = -\vec{\nabla}\cdot(\rho\vec{u}\vec{u}) - \nabla p + \vec{F}.
\end{equation}
This however is not yet the Navier-Stokes equation, in fact this is the Euler equation~\cite{batchelorIntroductionFluidDynamics1967}.
In the derivation there is one term missing, because the momentum of a fluid can also change due to internal friction, or put simply, due to viscosity.
These viscous contributions are collected in the viscous stress tensor $\hat{\sigma}$ as
\begin{equation}\label{eq:stress_tens}
    \hat{\sigma}_{ij} = \mu\left(\frac{\partial u_i}{\partial x_j} + \frac{\partial u_j}{\partial x_i}\right) + \lambda\delta_{ij}\frac{\partial u_k}{\partial x_k},
\end{equation}
This tensor introduces a lot of the complexity of the Navier-Stokes equation as in principle neither $\mu$ nor $\lambda$ need to be constants and in fact they are tensors themselves.

For the sake of clarity however here it is assumed that both $\mu$ and $\lambda$ are constants.
Inclusion of pressure to $\hat{\sigma}$ yields the total stress tensor $\tilde{\sigma}$. 
Due to symmetries of the problem there are various constraints on this tensor.
A fluid element can be rotated similar to a solid body, but it can also be stretched. 
In contrast to a solid body a fluid element is allowed to deform.
Working out these constraints we find that the pressure has to be a diagonal term and thus
\begin{equation}\label{eq:total_stress}
    \tilde{\sigma}_{ij} = - p \delta_{ij} + \mu\left(\frac{\partial u_i}{\partial x_j} + \frac{\partial u_j}{\partial x_i}\right) + \lambda\delta_{ij}\frac{\partial u_k}{\partial x_k}.
\end{equation}
Now to get from the Euler equation (Eq.~(\ref{eq:navier_stokes_2})) to the Navier-Stoke we add the viscous stress tensor and rearrange terms,
\begin{equation}\label{eq:navier_stokes_3}
    \partial_t(\rho\vec{u}) + \vec{\nabla}\cdot(\rho\vec{u}\vec{u}) = - \vec{\nabla} p + \vec{\nabla}\left[\mu(\vec{\nabla}\vec{u} + \vec{\nabla}\vec{u}^T) + \left(\eta-\frac{2\mu}{3}\right)\vec{\nabla}\cdot\vec{u}\right] + \vec{F},
\end{equation}
where we have used the differential form of Eq.~(\ref{eq:total_stress}) and $\lambda = \eta - 2/3\mu$.

@@

Inserting this constraint Eq.~(\ref{eq:incomp}) into Eq.~(\ref{eq:navier_stokes_3}) yields therefore the Navier-Stokes equation for an incompressible fluid as
\begin{equation}\label{eq:navier_stokes_4}
    \rho(\partial_t\vec{u} + \vec{\nabla}\cdot\vec{u}\vec{u}) = - \vec{\nabla} p + \mu\Delta\vec{u} + \vec{F},
\end{equation}
where the symbol $\Delta$ denotes the laplace operator ($\Delta f = \vec{\nabla}\cdot\vec{\nabla} f$).
The left hand side can be rewritten using the material derivative
\begin{equation}\label{eq:mat_dev}
    D_t = \partial_t + \vec{u}\cdot\vec{\nabla}, 
\end{equation}
which is a convenient measure for following a fluid element instead of a single point.
Inserting this, Eq.~(\ref{eq:navier_stokes_4}) can be written as
\begin{equation}\label{eq:navier_stokes_fin}
    \rho D_t \vec{u} =  - \vec{\nabla} p + \mu\Delta\vec{u} + \vec{F}.
\end{equation}
Eqs.~(\ref{eq:cont_3}, \ref{eq:navier_stokes_4}) cannot be solved because there are too few equations for too many unknowns.
So far there are five unknowns ($\rho, \mathbf{u}, p$) for four equations.
The clever way out of this misery is to use an equation of state that relates pressure, density, temperature, energy and entropy.
Especially the relation between pressure, density and temperature helps in closing the system of equations.
To define an equation of state is far from trivial.
It requires a thermodynamically consistent approach that is based on observations or theoretical assumptions.
The equation of state for an ideal gas was born from theoretical assumptions, e.g. non-interacting gas molecules, and reads~\cite{schwablStatisticalMechanics2006}
\begin{equation}\label{eq:eq_of_state}
    p = \rho R T,
\end{equation}
where $R$ is the specific gas constant and $T$ is the temperature.
Having a well defined state in the thermodynamical sense allows to relate pressure and density and therefore to close the system of equations.
The equation of state for an ideal gas, see Eq.~(\ref{eq:eq_of_state}), will be revisited in Chapter~\ref{chapter:method} where we discuss the lattice Boltzmann method.

The Navier-Stokes equation, see Eqs.~(\ref{eq:cont_3}, \ref{eq:navier_stokes_4}), offers a general framework to deal with fluid dynamics problems.
In the case of an incompressible non turbulent flow we have a solid understanding of the velocity field.
For compressible flows with developed turbulence on the other hand we still miss fundamental knowledge, as mentioned in Chapter~\ref{chapter:intro}.
This missing information is the reason that the automotive as well as aviation industry invests heavily in both real world experiments and CFD. 
For the concrete example of thin film flows, however, there is no need to ``solve'' the Navier-Stokes equation.
That is why in the next two sections different theories that build upon the Navier-Stokes are introduced.
Both of these theories using a scaling argument to severely simplify the resulting system of equations.

\section{Shallow Water Theory}
\label{sec:theory_shallow_water}
Solving the Navier-Stokes equation for the problems discussed in Chapters~\ref{chapter:fourth_paper}-\ref{chapter:third_paper} is not needed.
Looking at Figs.~\ref{fig:examples_intro}-\ref{fig:morph_transition} one can identify that the height or thickness of the fluid layer is much smaller than the horizontal extension of the substrate.
This statement together with some more approximations allows to derive another system from the Navier-Stokes equation and the continuity equation which is called the shallow water or Saint-Venant equations~\cite{bTheorieMouvementNonpermanent1871}.

@@

Third, the horizontal velocity of the fluid has to vanish at the ``solid'' ground, or in terms of Fig.~\ref{fig:shallow_water_drawing} looking down from the free surface at $-h(x,y)$.
Although there is a viscosity associated with the fluid, friction with the bottom topography is the dominating dissipation channel.
This kind of boundary condition has been introduced in Chapter~\ref{chapter:intro} is called \textbf{no-slip} boundary condition~\cite{richardsonNoslipBoundaryCondition1973, degennesFluidWallSlippage2002, salmonLatticeBoltzmannMethod1999}.
While this may seem counterintuitive because fluid has no problem to flow over yet dry surfaces, modelling wetting or the wet-dry transition is a complex task, as seen later in this chapter and in Chapter~\ref{chapter:first_paper}.
After all, a theoretical model which is defined as a set of partial differential equations cannot account for the true complexity of nature~\cite{scrivenPhysicsApplicationsDIP1988}.
We come back to the idea of the no-slip boundary condition in Sec.~\ref{sec:thin_films}.

@@

To close this section we quickly collect the equations that form the shallow water system, see Eqs.~(\ref{eq:shallow_water_cont_true}, \ref{eq:sw_mom_final}).
The generalization of these equations towards a second horizontal dimension is straightforward and can be written as~\cite{salmonLatticeBoltzmannMethod1999, dellarNonhydrodynamicModesPriori2002, thommesLatticeBoltzmannMethods2007}
\begin{align}    
        \partial_t h &+ \nabla \cdot (h\mathbf{u}) = 0, \label{eq:shallow_water_01}\\
        \partial_t (h\mathbf{u}) + \nabla&\cdot (p \mathbb{1} + h\mathbf{u}\mathbf{u} - S) = 0, \label{eq:shallow_water_02} 
\end{align}
where $\mathbf{u} = (u_x, u_y)$ and the only unknown $\mathcal{S}$ is a yet undefined source term. 
The system Eqs.~(\ref{eq:shallow_water_01}-\ref{eq:shallow_water_02}) concludes the shallow water model. 
In Chapter~\ref{chapter:first_paper} we show how this translates into a numerical solver.

@@

In the shallow water section, we quickly reduced the problem to two dimensions (one horizontal and one vertical).
Here we keep the second horizontal dimension ($\tilde{y}, \tilde{u}_y$), because most numerical experiments in Chapters~\ref{chapter:fourth_paper}-\ref{chapter:third_paper} are computed in a two dimensional domain.

@@

Now that the horizontal velocities are known we can simply integration the continuity equation Eq.~(\ref{eq:incomp})
\begin{equation}\label{eq:cont_thin}
    \int_0^h(\partial_x u_x + \partial_y u_y + \partial_z u_z) = 0,
\end{equation}
making use of the Leibniz rule one more time and showing only the $x$ component
\begin{equation}
    \int_0^h \partial_x u_x(z) \diff z = \partial_x\int_0^h u_x\diff z + \partial_x 0 u_x(0) - \partial_x h u_x(h) = \partial_x Q_x - \partial_x h u_x(h),  
\end{equation}
where Eq.~(\ref{eq:discharge_thin_x}) was used.
Inserting the above equation for the $x$ and $y$-component into the Eq.~(\ref{eq:cont_thin}) and performing the integration of $\partial_z u_z$ yields
\begin{equation}\label{eq:thin_dummy}
    \partial_x Q_x - (\partial_x h) u_x(h) + \partial_y Q_y - (\partial_y h) u_y(h) + u_z(h) - u_z(0) = 0.
\end{equation}
Since the no-slip boundary condition holds for all velocity components we set $u_z(0) = 0$.
The upper bound $u_z(h)$ can be computed using the material derivative, see Eq.~(\ref{eq:mat_dev}). 
We apply the same transformation as in Eq.~(\ref{eq:shallow_water_height}) with $f(\mathbf{x},t) = h(\mathbf{x},t) - z$, which is $0$ at the interface yields
\begin{equation}\label{eq:mat_uz_h}
    D_t f = \partial_t h + u_x\partial_x h + u_y\partial_y h - u_z = 0.
\end{equation}
Inserting the found result for $u(h)$ into Eq.~(\ref{eq:thin_dummy}) we get
\begin{equation}\label{eq:thin_obsc}
    \partial_x Q_x + \partial_y Q_y + \partial_t h = 0,
\end{equation}
which is almost the same as Eq.~(\ref{eq:simple_thin_film}).
Inserting $Q_x$ and $Q_y$, Eqs.~(\ref{eq:discharge_thin_x}) and (\ref{eq:discharge_thin_y}), we have
\begin{equation}
    \partial_t h(\mathbf{x},t) = \nabla\cdot\left(\frac{h^3}{3\mu}\nabla p + \frac{h^2}{2\mu}\tau\right),
\end{equation}
which compared to Eq.~(\ref{eq:simple_thin_film}) includes the shear contribution $\tau$.
Assuming the interface is not accelerated (e.g. in case of dewetting) the shear term can be neglected and the resulting equation is 
\begin{equation}\label{eq:thin_final}
    \partial_t h(\mathbf{x},t) = \nabla\cdot\left(\frac{h^3}{3\mu}\nabla p\right),
\end{equation}
which completes the task to derive the thin film equation from the Navier-Stokes equation. 
To sum up the approach in one sentence, we non dimensionalized the variables, collected terms in orders of $\varepsilon$, used our boundary conditions and performed an integration along the vertical dimension to average out the vertical dynamics.
Of course this derivation comes short if there are additional forces to consider such as thermal fluctuations, that said in Chapter~\ref{chapter:second_paper} we discuss the effect of thermal fluctuations on dewetting and show how the thin film equation has to be modified.

Because the thin film equation is the main theoretical framework for the following Chapters it is instructive to derive some more phenomenology of Eq.~(\ref{eq:thin_final}).
One helpful tool is the so called linear stability analysis. 
The idea is to slightly perturb the system, Eq.~(\ref{eq:thin_final}), and derive its response to the pertubation.

@@

An example of a dispersion relation which follows the above equation is shown in Fig.~\ref{fig:dispertion_1}.
In this figure we can identify two relevant wave numbers, on the one hand there is $q_0$ which is the fastest growing mode indicated by a dotted line and there is $q_m = \sqrt{2}q_0$ which is the largest wave number that is not damped.
Wavemodes larger than $q_m$ are therefore damped out.
A characteristic time scale of the problem is given as
\begin{equation}
    t_0 = \frac{3\mu}{\gamma h_0^3 q_0^4},
\end{equation}
with is used to normalize the time in Fig.~\ref{fig:dispertion_1}.
Because the concept of linear stability is such an important one, we encounter it again in Chapter~\ref{chapter:first_paper} see Fig.~\ref{fig:RTI} as well as in Chapter~\ref{chapter:second_paper} see Figs.~\ref{fig:theory_simulation_structure_factor}, \ref{fig:spectral_analysis_deter_sine8}, \ref{fig:spectral_analysis_stoch_sine8}, \ref{fig:square_wave_both}.
Chapter~\ref{chapter:second_paper} focuses only on the addition of a fluctuating term and the so called stochastic thin film equation.
The new term due to thermal fluctuations offers some interesting features on both the structure factor as well as the dispersion.

@@

The theoretical starting point of this Chapter is the Navier-Stokes equation.
This equation relates the change of momentum of the fluid to various effect, e.g. viscous contributions. 
As of today, nobody has solved the Millennium problem and showed that there exist a unique and smooth solution to this equation.
A well defined solution would have the benefit that even complex flow fields could be computed using analytical methods, such as pen and paper.
What we can do, is computing approximative results of this equation using numerical techniques.
This is in fact the idea of the following Chapter.
Mathematically the incompressible Navier-Stokes equation is a second order partial differential vector equation. 
It can be considered hyperbolic if the flow is advection dominated or parabolic if it is diffusion dominated.
The Reynolds number is the one dimensionless number that helps to distinguish between these two regimes.

>>>>

\section{The Navier-Stokes equation}
\label{sec:navier_stokes_sec}
Mathematically the Navier-Stokes equation can be derived from one definition and two constraints.
The two constraints are mass and momentum conservation.
Mass conservation simply states that mass cannot be lost or created.
The same argument holds for the momentum.
By Noethers theorem every conserved quantity yields a symmetry and vice versa.
In this case conservation of momentum allows us to use the constraints on the solutions that translational symmetry requires.
Put simply a sliding drop experiment has to yield the same result in any laboratory on earths surface, independent of the exact location of the laboratory.
The last ingredient for the derivation is the definition of a fluid.
A fluid is ``\textit{a substance that flows and is not solid}`` according to the Cambridge Dictionary.
This definition can be refined, because a fluid is either a gas or a liquid that continuously deforms under shear stress, at least in the case of a Newtonian fluid.
And as such there is a linear relation between shear rate and shear velocity for a Newtonian fluid.
Clearly there are other liquids (which encloses granular systems to some extent) that exhibit different behaviour, for example toothpaste.
As long as there is no force applied to the tube the toothpaste does not flow.
Even the acceleration due gravity when pointing the tube towards the ground does not exert enough force to generate a flow.
In contrast to Newtonian liquids those liquids have a different response function to shear, they can be addressed as \textit{power law} fluids.
While the phenomenology of these fluids, e.g. shear thinning or thickening, is very interesting and important in many industrial applications it is beyond the scope of this work. 
That said, in Chap.~\ref{chapter:second_paper} an extension towards more complex flows is presented. 
Instead of however changing the shear response we model thermal fluctuations to approximate the stochastic thin film equation (STF).
In fact, the step to derive an extension to our approach are nicely summarized in Chap.~\ref{chapter:second_paper}. 
It should serve as a guideline for further features, such as viscoelasticity for non-Newtonian liquids~\cite{bouchutNewModelShallow2013}.

@@

For the momentum equation the same line of arguments is applied, however momentum can change for different reasons.
Momentum can for example be advected with the flow from or into the volume $V_0$.
Pressure on the other hand can be a source of momentum change and of course forces that are applied to the fluid are as well sources~\cite{krugerLatticeBoltzmannMethod2017}.
Collecting these terms and writing down an equation for the momentum we have,
\begin{equation}\label{eq:navier_stokes_1}
    \partial_t \int_{V_0}\rho\vec{u}\diff V = -\oint_{\partial V_0}(\rho\vec{u}\vec{u})\cdot\diff\vec{A} - \oint_{\partial V_0} p\diff\vec{A} + \int_{V_0} \vec{F}\diff V, 
\end{equation}
where $\vec{u}\vec{u}$ is used to denote the outer product
\begin{equation}
    \vec{u}\vec{u} = \begin{bmatrix}
    u_x u_x & u_x u_y & u_x u_z \\
    u_y u_x & u_y u_y & u_y u_z \\
    u_z u_x & u_z u_y & u_z u_z
    \end{bmatrix} .
\end{equation}
Using the divergence theorem again to reformulate everything in terms of volume integrals and dropping the integration the above equation becomes
\begin{equation}\label{eq:navier_stokes_2}
    \partial_t(\rho\vec{u}) = -\vec{\nabla}\cdot(\rho\vec{u}\vec{u}) - \vec{\nabla} p + \vec{F}.
\end{equation}
This however is not yet the Navier-Stokes equation, in fact this is the Euler equation~\cite{batchelorIntroductionFluidDynamics1967}.
In the derivation there is one term missing, because the momentum of a fluid can also change due to internal friction, or put simply, due to viscosity.
These viscous contributions are collected in the viscous stress tensor $\hat{\sigma}$ as
\begin{equation}\label{eq:stress_tens}
    \hat{\sigma}_{\alpha\beta} = \mu\left(\frac{\partial u_{\alpha}}{\partial x_{\beta}} + \frac{\partial u_{\beta}}{\partial x_{\alpha}}\right) + \lambda\delta_{\alpha\beta}\frac{\partial u_{\gamma}}{\partial x_{\gamma}},
\end{equation}
This tensor introduces a lot of the complexity of the Navier-Stokes equation as in principle neither $\mu$ nor $\lambda$ need to be constants and in fact they are tensors themselves.

For the sake of clarity however here it is assumed that both $\mu$ and $\lambda$ are constants.
Inclusion of pressure to $\hat{\sigma}$ yields the total stress tensor $\tilde{\sigma}$. 
Due to symmetries of the problem there are various constraints on this tensor.
A fluid element can be rotated similar to a solid body, but it can also be stretched. 
In contrast to a solid body a fluid element is allowed to deform.
Working out these constraints we find that the pressure has to be a diagonal term and thus
\begin{equation}\label{eq:total_stress}
    \tilde{\sigma}_{\alpha\beta} = - p \delta_{\alpha\beta} + \mu\left(\frac{\partial u_{\alpha}}{\partial x_{\beta}} + \frac{\partial u_{\beta}}{\partial x_{\alpha}}\right) + \lambda\delta_{\alpha\beta}\frac{\partial u_{\gamma}}{\partial x_{\gamma}}.
\end{equation}
Now to get from the Euler equation (Eq.~(\ref{eq:navier_stokes_2})) to the Navier-Stoke we add the viscous stress tensor and rearrange terms,
\begin{equation}\label{eq:navier_stokes_3}
    \partial_t(\rho\vec{u}) + \vec{\nabla}\cdot(\rho\vec{u}\vec{u}) = - \vec{\nabla} p + \vec{\nabla}\left[\mu(\vec{\nabla}\vec{u} + \vec{\nabla}\vec{u}^T) + \left(\eta-\frac{2\mu}{3}\right)\vec{\nabla}\cdot\vec{u}\right] + \vec{F},
\end{equation}
where we have used the differential form of Eq.~(\ref{eq:total_stress}) and $\lambda = \eta - 2/3\mu$.

@@

Inserting this constraint Eq.~(\ref{eq:incomp}) into Eq.~(\ref{eq:navier_stokes_3}) yields therefore the Navier-Stokes equation for an incompressible fluid as
\begin{equation}\label{eq:navier_stokes_4}
    \rho(\partial_t\vec{u} + \vec{\nabla}\cdot\vec{u}\vec{u}) = - \vec{\nabla} p + \mu\Delta\vec{u} + \vec{F},
\end{equation}
where the symbol $\Delta$ denotes the laplace operator ($\Delta f = \vec{\nabla}\cdot\vec{\nabla} f$).
The left hand side can be rewritten using the material derivative
\begin{equation}\label{eq:mat_dev}
    D_t = \partial_t + \vec{u}\cdot\vec{\nabla}, 
\end{equation}
which is a convenient measure for following a fluid element instead of a single point.
Inserting this, Eq.~(\ref{eq:navier_stokes_4}) can be written as
\begin{equation}\label{eq:navier_stokes_fin}
    \rho D_t \vec{u} =  - \vec{\nabla} p + \mu\Delta\vec{u} + \vec{F}.
\end{equation}
Eqs.~(\ref{eq:cont_3}, \ref{eq:navier_stokes_4}) cannot be solved because there are too few equations for too many unknowns.
So far there are five unknowns ($\rho, \mathbf{u}, p$) for four equations.
The clever way out of this misery is to use an equation of state that relates pressure, density, temperature, energy and entropy.
Especially the relation between pressure, density and temperature helps in closing the system of equations.
To define an equation of state is far from trivial.
It requires a thermodynamically consistent approach that is based on observations or theoretical assumptions.
The equation of state for an ideal gas was born from theoretical assumptions, e.g. non-interacting gas molecules, and reads~\cite{schwablStatisticalMechanics2006}
\begin{equation}\label{eq:eq_of_state}
    p = \rho R T,
\end{equation}
where $R$ is the specific gas constant and $T$ is the temperature.
Having a well defined state in the thermodynamical sense allows to relate pressure and density and therefore to close the system of equations.
The equation of state for an ideal gas, see Eq.~(\ref{eq:eq_of_state}), will be revisited in Chap.~\ref{chapter:method} where we discuss the lattice Boltzmann method.

The Navier-Stokes equation, see Eqs.~(\ref{eq:cont_3}, \ref{eq:navier_stokes_4}), offers a general framework to deal with fluid dynamics problems.
In the case of an incompressible non turbulent flow we have a solid understanding of the velocity field.
For compressible flows with developed turbulence on the other hand we still miss fundamental knowledge, as mentioned in Chap.~\ref{chapter:intro}.
This missing information is the reason that the automotive as well as aviation industry invests heavily in both real world experiments and CFD. 
For the concrete example of thin film flows, however, there is no need to ``solve'' the Navier-Stokes equation.
That is why in the next two sections different theories that build upon the Navier-Stokes are introduced.
Both of these theories using a scaling argument to severely simplify the resulting system of equations.

\section{Shallow Water Theory}
\label{sec:theory_shallow_water}
Solving the Navier-Stokes equation for the problems discussed in Chaps.~\ref{chapter:fourth_paper}-\ref{chapter:third_paper} is not needed.
Looking at Figs.~\ref{fig:examples_intro}-\ref{fig:morph_transition} one can identify that the height or thickness of the fluid layer is much smaller than the horizontal extension of the substrate.
This statement together with some more approximations allows to derive another system from the Navier-Stokes equation and the continuity equation which is called the shallow water or Saint-Venant equations~\cite{bTheorieMouvementNonpermanent1871}.

@@

Third, the horizontal velocity of the fluid has to vanish at the ``solid'' ground, or in terms of Fig.~\ref{fig:shallow_water_drawing} looking down from the free surface at $-h(x,y)$.
Although there is a viscosity associated with the fluid, friction with the bottom topography is the dominating dissipation channel.
This kind of boundary condition has been introduced in Chap.~\ref{chapter:intro} is called \textbf{no-slip} boundary condition~\cite{richardsonNoslipBoundaryCondition1973, degennesFluidWallSlippage2002, salmonLatticeBoltzmannMethod1999}.
While this may seem counterintuitive because fluid has no problem to flow over yet dry surfaces, modelling wetting or the wet-dry transition is a complex task, as seen later in this chapter and in Chap.~\ref{chapter:first_paper}.
After all, a theoretical model which is defined as a set of partial differential equations cannot account for the true complexity of nature~\cite{scrivenPhysicsApplicationsDIP1988}.
We come back to the idea of the no-slip boundary condition in Sec.~\ref{sec:thin_films}.

@@

To close this section we quickly collect the equations that form the shallow water system, see Eqs.~(\ref{eq:shallow_water_cont_true}, \ref{eq:sw_mom_final}).
The generalization of these equations towards a second horizontal dimension is straightforward and can be written as~\cite{salmonLatticeBoltzmannMethod1999, dellarNonhydrodynamicModesPriori2002, thommesLatticeBoltzmannMethods2007}
\begin{align}    
        \partial_t h &+ \nabla \cdot (h\mathbf{u}) = 0, \label{eq:shallow_water_01}\\
        \partial_t (h\mathbf{u}) + \nabla&\cdot (p \mathbb{1} + h\mathbf{u}\mathbf{u} - S) = 0, \label{eq:shallow_water_02} 
\end{align}
where $\mathbf{u} = (u_x, u_y)$ and the only unknown $\mathcal{S}$ is a yet undefined source term. 
The system Eqs.~(\ref{eq:shallow_water_01}-\ref{eq:shallow_water_02}) concludes the shallow water model. 
In Chap.~\ref{chapter:first_paper} we show how this translates into a numerical solver.

@@

In the shallow water section, we quickly reduced the problem to two dimensions (one horizontal and one vertical).
Here we keep the second horizontal dimension ($\tilde{y}, \tilde{u}_y$), because most numerical experiments in Chaps.~\ref{chapter:fourth_paper}-\ref{chapter:third_paper} are computed in a two dimensional domain.

@@

Now that the horizontal velocities are known we can simply integration the continuity equation Eq.~(\ref{eq:incomp})
\begin{equation}\label{eq:cont_thin}
    \int_0^h(\partial_x u_x + \partial_y u_y + \partial_z u_z) = 0,
\end{equation}
making use of the Leibniz rule one more time and showing only the $x$ component
\begin{equation}
    \int_0^h \partial_x u_x(z) \diff z = \partial_x\int_0^h u_x\diff z + \partial_x 0 u_x(0) - \partial_x h u_x(h) = \partial_x Q_x - \partial_x h u_x(h),  
\end{equation}
where Eq.~(\ref{eq:discharge_thin_x}) was used.
Inserting the above equation for the $x$ and $y$-component into the Eq.~(\ref{eq:cont_thin}) and performing the integration of $\partial_z u_z$ yields
\begin{equation}\label{eq:thin_dummy}
    \partial_x Q_x - (\partial_x h) u_x(h) + \partial_y Q_y - (\partial_y h) u_y(h) + u_z(h) - u_z(0) = 0.
\end{equation}
Since the no-slip boundary condition holds for all velocity components we set $u_z(0) = 0$.
The upper bound $u_z(h)$ can be computed using the material derivative, see Eq.~(\ref{eq:mat_dev}). 
We apply the same transformation as in Eq.~(\ref{eq:shallow_water_height}) with $f(\mathbf{x},t) = h(\mathbf{x},t) - z$, which is $0$ at the interface yields
\begin{equation}\label{eq:mat_uz_h}
    D_t f = \partial_t h + u_x\partial_x h + u_y\partial_y h - u_z = 0.
\end{equation}
Inserting the found result for $u(h)$ into Eq.~(\ref{eq:thin_dummy}) we get
\begin{equation}\label{eq:thin_obsc}
    \partial_x Q_x + \partial_y Q_y + \partial_t h = 0,
\end{equation}
which is almost the same as Eq.~(\ref{eq:simple_thin_film}).
Inserting $Q_x$ and $Q_y$, Eqs.~(\ref{eq:discharge_thin_x}) and (\ref{eq:discharge_thin_y}), we have
\begin{equation}
    \partial_t h(\mathbf{x},t) = \nabla\cdot\left(\frac{h^3}{3\mu}\nabla p + \frac{h^2}{2\mu}\tau\right),
\end{equation}
which compared to Eq.~(\ref{eq:simple_thin_film}) includes the shear contribution $\tau$.
Assuming the interface is not accelerated (e.g. in case of dewetting) the shear term can be neglected and the resulting equation is 
\begin{equation}\label{eq:thin_final}
    \partial_t h(\mathbf{x},t) = \nabla\cdot\left(\frac{h^3}{3\mu}\nabla p\right),
\end{equation}
which completes the task to derive the thin film equation from the Navier-Stokes equation. 
To sum up the approach in one sentence, we non dimensionalized the variables, collected terms in orders of $\varepsilon$, used our boundary conditions and performed an integration along the vertical dimension to average out the vertical dynamics.
Of course this derivation comes short if there are additional forces to consider such as thermal fluctuations, that said in Chap.~\ref{chapter:second_paper} we discuss the effect of thermal fluctuations on dewetting and show how the thin film equation has to be modified.

Because the thin film equation is the main theoretical framework for the following Chaps. it is instructive to derive some more phenomenology of Eq.~(\ref{eq:thin_final}).
One helpful tool is the so called linear stability analysis. 
The idea is to slightly perturb the system, Eq.~(\ref{eq:thin_final}), and derive its response to the pertubation.

@@

An example of a dispersion relation which follows the above equation is shown in Fig.~\ref{fig:dispertion_1}.
In this figure we can identify two relevant wave numbers, on the one hand there is $q_0$ which is the fastest growing mode indicated by a dotted line and there is $q_m = \sqrt{2}q_0$ which is the largest wave number that is not damped.
Wavemodes larger than $q_m$ are therefore damped out.
A characteristic time scale of the problem is given as
\begin{equation}
    t_0 = \frac{3\mu}{\gamma h_0^3 q_0^4},
\end{equation}
with is used to normalize the time in Fig.~\ref{fig:dispertion_1}.
Because the concept of linear stability is such an important one, we encounter it again in Chap.~\ref{chapter:first_paper} see Fig.~\ref{fig:RTI} as well as in Chap.~\ref{chapter:second_paper} see Figs.~\ref{fig:theory_simulation_structure_factor}, \ref{fig:spectral_analysis_deter_sine8}, \ref{fig:spectral_analysis_stoch_sine8}, \ref{fig:square_wave_both}.
Chap.~\ref{chapter:second_paper} focuses only on the addition of a fluctuating term and the so called stochastic thin film equation.
The new term due to thermal fluctuations offers some interesting features on both the structure factor as well as the dispersion.

@@

The theoretical starting point of this chapter is the Navier-Stokes equation.
This equation relates the change of momentum of the fluid to various effect, e.g. viscous contributions. 
As of today, nobody has solved the Millennium problem and showed that there exist a unique and smooth solution to this equation.
A well defined solution would have the benefit that even complex flow fields could be computed using analytical methods, such as pen and paper.
What we can do, is computing approximative results of this equation using numerical techniques.
This is in fact the idea of the following chapter.
Mathematically the incompressible Navier-Stokes equation is a second order partial differential vector equation. 
It can be considered hyperbolic if the flow is advection dominated or parabolic if it is diffusion dominated.
The Reynolds number is the one dimensionless number that helps to distinguish between these two regimes.
